---
title: Making Custom Census Estimates using Public Use Microdata Sample data and srvyr
author:
date: "2018-11-06"
slug: pums
draft: true
tags:
  - rstats
  - census
  - srvyr
---

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(
  cache = TRUE, warning = FALSE, message = FALSE, collapse = TRUE,
  echo = TRUE, dpi = 180, cache.lazy = FALSE,
  fig.retina = 2
)

knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
options(width = 77, dplyr.width = 77)

## load all libraries for post here
library(tidyverse)
library(janitor)
library(ggridges)
library(scales)
library(viridis)

```

```{r read in}
library(tidyverse)
library(janitor)

nj_pums <- read_csv(
  #file = "https://mattherman.info/csv/nj_pums.csv",
  file = "/Users/matthewherman/Downloads/nj_pums.csv",
  col_types = cols(.default = "c")) %>%
  select(
    PUMA,
    AGEP,    # age
    SEX,     # sex
    RAC1P,   # race
    HISP,    # detailed hispanic origin
    POVPIP,  # income to poverty ratio
    WAGP,    # wage or salary income past 12 months
    WKHP,    # hours worked per week
    JWMNP,   # travel time to work in minutes
    JWTR,    # means of transportation to work
    PWGTP,   # survery weight
    PWGTP1:PWGTP80  # survey replicate weight
  ) %>% 
  clean_names()  # make all var names lowercase
```

```{r process}
nj_pums_processed <- nj_pums %>% 
  mutate_at(vars(-puma), as.integer) %>%  # convert to integer variables
  filter(!is.na(jwmnp) & wkhp >= 30 & wagp > 10000) %>% # remove non workers and those who work at home
  mutate(
    sex = if_else(sex == 1, "male", "female"),
    
    race = case_when(
      rac1p == 1 & hisp == 1      ~ "white",
      rac1p == 2 & hisp == 1      ~ "black",
      rac1p %in% 6:7 & hisp == 1  ~ "asian",
      rac1p %in% 3:5 & hisp == 1  ~ "native",
      rac1p %in% 8:9 & hisp == 1  ~ "two_other",
      hisp != 1                   ~ "latino",
    ),
    
    commute_method = case_when(
      jwtr %in% c(1, 7, 8) ~ "car",  # combining car, taxi motorcycle
      jwtr == 2            ~ "bus",
      jwtr %in% 3:4        ~ "subway",
      jwtr == 5            ~ "train",
      jwtr == 6            ~ "ferry",
      jwtr == 9            ~ "bicycle",
      jwtr == 10           ~ "walk",
      jwtr == 12           ~ "other"
    ),
    
    commute_public = if_else(
      commute_method %in% c("bus", "subway", "train", "ferry"),
      1L, 0L
      ),
    
    wage_log = log10(wagp)
  ) %>% 
  rename(
    age = agep,
    wage = wagp,
    work_hours = wkhp,
    work_travel_time = jwmnp
    ) %>% 
  select(-(rac1p:hisp), -jwtr)
```

```{r commute time ridge plot}
nj_pums_processed %>% 
  filter(work_travel_time <= 120 & commute_method != "other") %>%  
  mutate(
    commute_method = fct_reorder(str_to_title(commute_method), work_travel_time),
  ) %>%   
  select(work_travel_time, commute_method, pwgtp) %>% 
  uncount(weights = pwgtp) %>% 
  ggplot(
    aes(
      x = work_travel_time,
      y = commute_method,
      fill = ..x..
      )
    ) +
  stat_density_ridges(
    bandwidth = 6,
    geom = "density_ridges_gradient",
    gradient_lwd = 1.5
    ) +
  scale_fill_viridis(
    name = "Travel Time\n(minutes)",
    option = "C",
    breaks = seq(0, 120, 30)
    ) +
  scale_x_continuous(expand = c(0.01, 0), breaks = seq(0, 120, 30)) +
  scale_y_discrete(expand = c(0.01, 0)) +
  labs(
    x = "Daily travel time to work (minutes)",
    y = "",
    title = "How long do New Jerseyans spend commuting each day?",
    caption = "Source: 2017 American Community Survey PUMS"
  ) +
  theme_ridges(center_axis_labels = FALSE) +
  theme(
    legend.key.height = unit(15, "mm"),
    legend.key.width = unit(8, "mm"),
    plot.caption = element_text(size = 8, face = "italic")
    )
```

```{r}
nj_pums_processed %>% 
  mutate(work_travel_cat = cut(work_travel_time, breaks = (c(seq(0, 200, 15), Inf)))) %>% 
  select(work_travel_cat, wage, pwgtp) %>% 
  uncount(weights = pwgtp) %>% 
  ggplot(aes(x = work_travel_cat, y = wage)) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar)

```




ggplot(iris, aes(x = Sepal.Length, y = Species)) + geom_density_ridges()




nj_pums_processed %>% 
  count(commute_public, wt = pwgtp)

nj_pums_processed %>% 
  count(commute_means, wt = pwgtp, sort = TRUE)

nj_pums_processed %>%
  summarise_at(
    vars(wage),
    funs(mean, sd, median, max, min)
  )
  
nj_pums_processed %>%
  summarise_at(
    vars(work_travel_time),
    funs(mean, sd, median, max, min)
  )

hi %>% 
  filter(wage > 1000) %>% 
  ggplot(aes(x = work_travel_cat, y = wage)) +



hi<- nj_pums_processed %>% 





nj_pums_processed





```
