---
title: "Using SearchPanes extension for DT"
author:
date: "2020-05-17"
slug: dt_searchpanes
draft: true
tags:
  - rstats
  - dt
  - datatables
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  echo = TRUE,
  dpi = 180,
  fig.retina = 2
  )

options(width = 77, dplyr.width = 77)

library(tidyverse)
library(DT)
library(gapminder)
```

This is the first in a series of posts that describes small enhancements that you can add to your R Markdown documents and dashboards to give them a more "interactive feel" while not relying on Shiny for that interactivity. All of the interactions I will describe are executed on the client-side and can be implemented with a static, self-contained HTML file. This is a nice featured because it means you can host your documents on a simple static webhost like GitHub Pages or Netlify without relying on a backend. Additionally if you don't have access to an internal Shiny server at work (like me!) you can simply email these files around as attachments.

I'm not sure how many posts there will be in the series, but since I've recently been creating a number of these dashboard and sites at [work](https://www1.nyc.gov/site/acs/about/data-analysis.page), I hope to document many of the tips and tricks I'm learning along the way.

In this series, I'm going to use data from the [gapminder package](https://github.com/jennybc/gapminder), so if you don't have it installed:

```{r eval=FALSE}
install.packages("gapminder")
```

If you've created tables in R Markdown documents for HTML or Shiny, you've probably used the [DT package](https://rstudio.github.io/DT/) which is an R interface to the DataTables JavaScript library. I won't give an introduction to how to use DT, but the best place to start is the [DT website](https://rstudio.github.io/DT/), which has lots of documentation and examples.


Tip 1: Use SearchPanes for selecting rows in a DataTable

```{r}
library(tidyverse)
library(DT)
library(gapminder)

gapminder %>% 
  datatable()
  

```
