---
title: "Using the SearchPanes extension with DT"
author:
date: "2020-05-17"
slug: dt_searchpanes
draft: false
tags:
  - rstats
  - dt
  - datatables
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  echo = TRUE,
  dpi = 180,
  fig.retina = 2
  )

options(width = 77, dplyr.width = 77)

library(DT)
library(gapminder)
```

This is the first in a series of posts that describes small enhancements that you can add to your R Markdown documents and dashboards to give them a more "interactive feel" while not relying on Shiny for that interactivity. All of the interactions I will describe are executed on the client-side and can be implemented with a static, self-contained HTML file. This is a nice feature because it means you can host your documents using a simple static webhost like [GitHub Pages](https://pages.github.com/) or [Netlify](https://www.netlify.com/blog/2016/10/27/a-step-by-step-guide-deploying-a-static-site-or-single-page-app/). Additionally if you don't have access to an internal Shiny server at work (like me!) you can email these files around as attachments.

I'm not sure how many posts there will be in the series, but since I've recently been creating a number of these dashboard and sites at [work](https://www1.nyc.gov/site/acs/about/data-analysis.page), I hope to document many of the tips and tricks I'm learning along the way.

Here are the other posts in the series:

[**Tip 2: Order string dates by an invisible DataTable column**](/blog/dt_sort_secondary)

In this post, I'm going to use data from the [gapminder package](https://github.com/jennybc/gapminder), so if you don't have it installed:

```{r eval=FALSE}
install.packages("gapminder")
```

If you've created tables in R Markdown documents for HTML or Shiny, you've probably used the [DT package](https://rstudio.github.io/DT/) which is an R interface to the [DataTables JavaScript library](https://datatables.net/). I won't give an introduction to DT, but the best place to start is the [DT website](https://rstudio.github.io/DT/), which has lots of examples.

**Tip 1: Use SearchPanes to select rows in a DataTable**

If you want to do filtering or row selection, the standard DataTables search box is not very flexible. A new option is the SearchPanes extension that was [added to DataTables in early 2020](https://datatables.net/blog/2020-01-13) and quickly implemented in [DT as well](https://github.com/rstudio/DT/pull/756). It is [very well documented](https://datatables.net/extensions/searchpanes/) on the DataTables website.

The basic usage of SearchPanes allows you to filter multiple columns using multi-select (hold down <kbd>SHIFT</kbd>!) for each column. To get started, just add the `Select` and `SearchPanes` as extensions when you create a DataTable. You also need to insert `P` as a `dom` option to display the panes. Here's a minimal  example using gapminder data:

```{r}
library(DT)
library(gapminder)

gapminder %>% 
  datatable(
    extensions = c("SearchPanes", "Select"),
    options = list(dom = "Ptip")
    )
```
<br>
<br>
Another nice feature of SearchPanes is that it shows you the number of rows in each category. For example, at a quick glance we can see there are 12 rows for each country and 624 of the rows are African countries.

Depending on the layout of your final table and document, you may find the search pane taking up too much valuable real estate. A good way to use the SearchPanes feature in a smaller footprint is to hide the pane in a button. [Buttons](https://datatables.net/extensions/buttons/) is another extension for DataTables that is supported by DT.

In order to create a button for the search pane, add `"Buttons"` to `extensions`, set the buttons options to use `searchPanes` (note the camelCase here!), and replace `P` with `B` in the `dom` options.

```{r}
gapminder %>% 
  datatable(
    extensions = c("SearchPanes", "Select", "Buttons"),
    options = list(
      dom = "Btip",
      buttons = list("searchPanes")
      )
    )
```
<br>
<br>
Now, the search pane pops open when you click the button.

There is a ton more customization that is possible using the new SearchPanes extension, including [modifying the layout](https://datatables.net/extensions/searchpanes/layout), creating [custom filtering functions](https://datatables.net/extensions/searchpanes/customSearchFunctions), and more. But a few simple modifications I'll demonstrate are: 1) choosing which columns to search, 2) removing the SearchPane controls, and 3) changing the text of the SearchPanes button.  

1. To limit which columns are active for searching, add `searchPanes = list(show = FALSE), targets = 3:5)` to `columnDefs`. This hides columns 3, 4, and 5, and only displays search panes for the first 3 columns (JS uses 0-based indexing!). The SearchPanes default is to included all columns.

2. For a simple table, you might not need all of the SearchPanes controls, which allow for sorting and searching *within* the search pane. To remove the controls, add `list(searchPanes = list(controls = FALSE), targets = 0:2)` to the columnDefs option. We need to set this option for the indices of the columns we've chosen to add SearchPanes to.[^1]

[^1]: I had some trouble figuring out how to hide the controls when I was using SearchPanes within a button. Thank you so much to [shrektan](https://shrektan.com/) for [pointing me in the right direction](https://github.com/rstudio/DT/issues/804#issuecomment-620637155) for this option that doesn't seem to be documented.

3. "SearchPanes" isn't such a great label for the button. Let's call it "Filter Rows" instead, by setting `language = list(searchPanes = list(collapse = "Filter Rows"))`.[^2]

[^2]: Candidly, I need [shrektan's help](https://github.com/rstudio/DT/issues/804#issuecomment-620610263) for this one too!

While we're tidying things up, let's set better column names, format the numeric variables, get rid of the row numbers, center all the columns, and replace the pagination with a the [Scroller extension](https://datatables.net/extensions/scroller/).

```{r}
gapminder %>% 
  datatable(
    rownames = FALSE,
    colnames = c("Country", "Continent", "Year",
                 "Life Exp", "Population", "GDP per capita"),
    extensions = c("SearchPanes", "Select", "Buttons", "Scroller"),
    options = list(
      dom = "Btip",
      buttons = list("searchPanes"),
      language = list(searchPanes = list(collapse = "Filter Rows")),
      scrollY = 390,
      scroller = TRUE,
      columnDefs = list(
        list(searchPanes = list(show = FALSE), targets = 3:5),
        list(searchPanes = list(controls = FALSE), targets = 0:2),
        list(className = "dt-center", targets = 0:5)
        )
      )
    ) %>% 
  formatRound(4, 1) %>% 
  formatRound(5, 0) %>% 
  formatCurrency(6, digits = 0)
```
<br>

And there you have it, a nice table that makes it easy to view gapminder data from Oceania in 1972 and 1977!
